---
title: "Wikimedia Analytics Query Service (AQS) API in R"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

{waxer} is an R wrapper for the [Wikimedia Analytics Query Service (AQS)](https://wikitech.wikimedia.org/wiki/Analytics/Systems/AQS). This particular wrapper is for the `/metrics` endpoint of the [REST API](https://wikimedia.org/api/rest_v1/) which provides data and metrics around traffic, users, and content on Wikimedia sites.

__Author:__ Mikhail Popov (Wikimedia Foundation)<br/>
__License:__ [MIT](http://opensource.org/licenses/MIT)<br/>
__Status:__ Active

## Installation
    
To install the development version:

```R
# install.packages("remotes")
remotes::install_github("wikimedia/waxer@main")
```

To update:

```R
remotes::update_packages(packages = "waxer")
```

## Usage

```{r load, eval=FALSE}
library(waxer)
?waxer
```

## Additional Information

Please note that this project is released with a [Contributor Code of Conduct](CONDUCT.md). By participating in this project you agree to abide by its terms.

**Why "waxer"?** Well, I was looking at the initialism "WAQSR" (Wikimedia Analytics Query Service in R) and just really not liking how awkward it looked.
